---
title: Bioconductor-friendly bindings for spatial classes
author:
- name: Kevin Rue-Albrecht
  email: kevinrue67@gmail.com
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
package: spatula
---

```{r}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
set.seed(1000)
```

# Introduction

The `r CRANpkg("sp")` package provides a number of useful classes and methods for working with spatial data.
The `r Biocpkg("spatula")` package extends these classes to make them work more naturally with Bioconductor data structures.
This allows us to, for example, attach spatial information to column metadata of `SummarizedExperiment` objects
in order to relate samples to their spatial coordinates.
Importantly, `r Biocpkg("spatula")` only provides new methods for but does not modify the `r CRANpkg("sp")` classes, 
meaning that it is directly compatible with other packages that take instances of such classes as input.

# Storing in a `DataFrame`

It is straightforward to store, say, an instance of the `SpatialPoints` class as a column of a `DataFrame`.
In fact, this is possible to do without `r Biocpkg("spatula")`, but the package provides a nice print method for such objects:

```{r}
library(spatula)
coords <- matrix(rnorm(100), ncol=2)
points <- SpatialPoints(coords)
df <- DataFrame(location=I(points))
df
```

This implies that we can easily store spatial coordinates in the metadata fields of a `SummarizedExperiment`.
The most common application for this functionality is to track the spatial location of samples in the `colData`,
but one could also imagine applying this to `rowData` for features that have some spatial association.

```{r}
library(SummarizedExperiment)
se <- SummarizedExperiment(list(counts=matrix(rpois(1000, 1), ncol=50)))
se$location <- points
se
```

As one might expect, subsetting the `SummarizedExperiment` will also subset the `SpatialPoints` object embedded within:

```{r}
first10 <- se[,1:10]
first10$location
```

# Basic operations on `SpatialPoints`

For reasons unbeknownst to us, `r CRANpkg("sp")` decided not to implement a concatenation operator for their `SpatialPoints` classes.
So, here it is:

```{r}
coords2 <- matrix(rnorm(20), ncol=2)
points2 <- SpatialPoints(coords2)

combined <- c(points, points2)
length(combined)
head(combined)
```

We also provide utilities like `sort()` and `unique()` for `SpatialPoints` objects.
Comparisons between points are performed by comparing the vector of coordinates,
where the "larger" point is the one with the larger value in the first non-tied dimension.
Obviously, this only works for `SpatialPoints` objects with the same dimensions.

```{r}
head(sort(points))
```

The `match()` function identifies matching points in two different objects,
which is potentially useful for comparing annotations across different experiments.
Note that this is done using an exact comparison of the double-precision coordinates, so it literally has to be the exact same point - 
no tolerance is accepted (see `findOverlaps()` below if that is of interest).

```{r}
match(points2, combined)
```

# Basic operations on `SpatialPolygons`

Creating a `SpatialPolygons` object can be somewhat painful - 
one has to create a `Polygon`, and then a `Polygons`, and then finally a `SpatialPolygons` -
so we created a wrapper function that smoothes the path in the simplest case.
Here, each matrix represents the vertices of a single 2-dimensional polygon:

```{r}
# Four quadrants:
poly <- makeSpatialPolygons(
    rbind(c(0,0), c(0.5, 0), c(0.5, 0.5), c(0, 0.5)),
    rbind(c(1,1), c(0.5, 1), c(0.5, 0.5), c(1, 0.5)),
    rbind(c(0.5,0), c(1, 0), c(1, 0.5), c(0.5, 0.5)),
    rbind(c(0,1), c(0.5, 1), c(0.5, 0.5), c(0, 0.5))
)
```

Like before, we can store this in a `DataFrame` where it is printed with the coordinates of the centroids of each polygon.
In fact, this print method is prettier than that of the `SpatialPolygons` itself, 
which (as of time of writing) just shows the raw S4 structure.

```{r}
X <- DataFrame(polygon=I(poly))
X
```

We also provide methods for `c()`, `sort()`, `unique()` and `match()`, as described above.
Sorting is performed based on a comparison of the vector of coordinates for the cluster centroids;
if those are equal, then polygons are sorted by area.

```{r}
o <- order(poly)
X[o,,drop=FALSE]
```

# Finding overlaps

`r Biocpkg("spatula")` wraps the `sp::over()` method into the more familiar semantics of `findOverlaps()`.
Given `SpatialPoints` and `SpatialPolygons` objects, we can easily identify which points overlap each polygon:

```{r}
hits <- findOverlaps(poly, points)
hits
```

We use the same method to identify which polygons overlap each other, 
though this requires further installation of the `r CRANpkg("rgeos")` package:

```{r}
other.poly <- makeSpatialPolygons(
    rbind(c(0.1,0.1), c(0.4, 0.1), c(0.4, 0.4), c(0.1, 0.4))
)
findOverlaps(poly, other.poly)
```

# Session information

```{r}
sessionInfo()
```
